Adaptive Algorithms Employing Continuous Linear Functionals

### 2018-08-08 AutoAlgorithmLinearFunctionalDataVerC.tex
Polishing and correcting typos

\section{Introduction}
Adaptive algorithms determine the design and sample size needed to solve problems to the desired accuracy.  A priori upper bounds on some norm of the input function are not required, but there may be some underlying assumptions about the input function required for the adaptive algorithm to be successful.  Here we consider general linear problems where the series coefficients of the input function are used to obtain an approximate solution.  The algorithm proposed guarantees the accuracy of the approximate solution.  Moreover, our algorithm is shown to have a computational cost that is asymptotically optimal in comparison to the best possible algorithm.  The set of functions on which our adaptive algorithm is defined is a cone.

Let $\cf$ be a separable Hilbert space of input functions with orthonormal basis $\{u_i\}_{i=1}^\infty$, and let $\cg$ be a separable Hilbert space of outputs  with orthonormal basis $\{v_i\}_{i=1}^\infty$:  
\begin{subequations}\label{probDef}
\begin{gather}
f = \sum_{i=1}^\infty \hf_i u_i \in \cf, \qquad \norm[\cf]{f}=\bignorm[2]{(\hf_i\big)_{i=1}^\infty}, \\
g = \sum_{i=1}^\infty \hg_i v_i \in \cg, \qquad \norm[\cg]{g}=\bignorm[2]{(\hg_i\big)_{i=1}^\infty}.
\end{gather}
Let these bases be chosen by the singular value decomposition so that the linear solution operator, $S:\cf \to \cg$, satisfies
\begin{gather}
S(u_i) = \lambda_i v_i, \quad i \in \naturals, \qquad S(f) = \sum_{i=1}^n \lambda_i \hf_i v_i, \\
\lambda_1 \ge \lambda_2 \ge \cdots > 0, \quad \lim_{i \to \infty} \lambda_i = 0, \qquad
\norm[\cf \to \cg]{S} := \sup_{f \ne 0} \frac{\norm[\cg]{S(f)}}{\norm[\cf]{f}} = \lambda_1.
\end{gather}
\end{subequations}
This setting includes, for example, the recovery of functions, derivatives, and indefinite integrals.  We are interested in cases where the exact solution generally requires an infinite number of series coefficients, $\hf_i$, Thus, this setting is not relevant for evaluating the definite integral or any other linear functional, since in those cases $\lambda_2 = \lambda_3 = \cdots = 0$.

Let $\ch$ be any subset of $\cf$, and let $\ca(\ch)$ denote the set of deterministic algorithms that successfully approximate the solution to some error tolerance for all inputs in $\ch$:
\begin{multline} \label{adapErrCrit}
\ca(\ch) : = \left\{ \text{algorithms } A:\ch \times (0,\infty) \rightarrow \cg : 
\right . \\ \left .
\bignorm[\cg]{S(f) - A(f,\varepsilon)} \le \varepsilon \ \forall f \in \ch, \ \varepsilon > 0 
\right \}.
\end{multline}
Algorithms in $\ca(\ch)$ are allowed to sample adaptively any bounded, linear functionals
of the input function $f$.  They must require only a finite number of linear functionals for each input function and positive tolerance.  The definition of $\ch$ can be used in constructing algorithms in $\ca(\ch)$, but no other a priori knowledge about the input functions is available.  Following \cite{KunEtal19a} we call a problem \emph{solvable} for inputs $\ch$ if $\ca(\ch)$ is non-empty.

Our problem is not solvable for $\cf$, as can be demonstrated by contradiction. For any potential algorithm, we show that there is some $f \in \cf$, that looks like $0$ to the algorithm, but for which $S(f)$ is far from $S(0) = 0$.   Choose any $A \in \ca(\cf)$ and $\varepsilon > 0$, and let $L_1, \ldots, L_n$ be the linear functionals are used to compute $A(0,\varepsilon)$. Since $n$ is finite, there exists some nonzero $f \in \cf$ satisfying that $L_1(f) = \cdots = L_n(f) = 0$ with non-zero $S(f)$. This means that $A(cf,\varepsilon) = A(0,\varepsilon)$ for any real $c$.  Since $A$ successfully produces an approximation with error no greater than $\varepsilon$ for both $0$ and $cf$,
\begin{align*}
    \varepsilon &\ge \frac 12 \left[ \norm[\cg]{S(0) - A(0,\varepsilon)} + \norm[\cg]{S(cf) - A(cf,\varepsilon)}\right] \\
    & =  \frac 12 \left[ \norm[\cg]{0 - A(0,\varepsilon)} + \norm[\cg]{S(cf) - A(0,\varepsilon)}\right] \\
    & \ge  \frac {\abs{c}\norm[\cg]{S(f)}}2  \qquad \text{by the triangle inequality}.
\end{align*}
Since $S(f) \ne 0$, it is impossible for this inequality to hold for all real $c$.  Thus, $\ca(\cf)$ is empty, and our problem is not solvable for $\cf$.  However, it is solvable for well-chosen subsets of $\cf$.

The computational cost of an algorithm $A \in \ca(\ch)$ for $f \in \ch$ and error tolerance $\varepsilon$ is denoted $\cost(A,f,\varepsilon)$, and is defined as the number of linear functional values required to produce $A(f,\varepsilon)$.  By overloading the notation, we define for arbitrary algorithms $A \in \ca(\ch)$, 
\begin{equation*}
\cost(A,\ch,\varepsilon) : = \sup \{\cost(A,f,\varepsilon) : f \in \ch \} \qquad \forall \varepsilon > 0.
\end{equation*}
For unbounded sets, $\ch$, this cost may be infinite.  Therefore, it is also helpful to define the cost for input functions in $\ch \cap \cb_{\rho}$, where $\cb_{\rho} : = \{ f \in \cf : \norm[\cf]{f} \le \rho \}$ is the ball of radius $\rho$:
\begin{equation*}
\cost(A,\ch,\varepsilon,\rho) : = \sup \{\cost(A,f,\varepsilon) : f \in \ch \cap \cb_{\rho} \} \qquad \forall \rho > 0, \ \varepsilon > 0.
\end{equation*}
Finally, we define the complexity of the problem as computational cost of the best algorithm:
\begin{gather*}
\comp(\ca(\ch),\varepsilon) := \min_{A \in \ca(\ch)} \cost(A, \ch, \varepsilon), \\
\comp(\ca(\ch),\varepsilon,\rho) := \min_{A \in \ca(\ch)} \cost(A, \ch, \varepsilon, \rho).
\end{gather*}
Note that $\comp(\ca(\ch),\varepsilon,\rho) \ge \comp(\ca(\ch \cap \cb_{\rho}),\varepsilon)$.  In the former case, the algorithm is unaware  that the input function has norm no greater than $\rho$.

@@

To prove this, let $A^*$ be an arbitrary algorithm in $\ca(\cb_{\rho})$, and let $L_1, \ldots, L_n$ be the linear functionals chosen when evaluating this algorithm for the zero function with tolerance $\varepsilon$.  Thus, $A^*(0,\varepsilon)$ is some function of $(L_1(0) , \ldots, L_n(0)) = (0, \ldots, 0)$.  Let $f$ be a linear combination of $u_1, \ldots, u_{n+1}$ with norm $\rho$ satisfying  $L_1(f) = \cdots = L_n(f) = 0$, then $A^*(\pm f) = A^*(0)$, and
\begin{align*}
\varepsilon & \ge \max_{\pm} \norm[\cg]{S(\pm f) - A^*(\pm f)} =  \max_{\pm} \norm[\cg]{\pm S( f) - A^*(0)} \\
& \ge \frac 12 \left [ \norm[\cg]{S( f) - A^*(0)} + \norm[\cg]{- S( f) - A^*(0)}\right] \\
& \ge \norm[\cg]{S(f)} 
= \bignorm[2]{\big(\lambda_i\hf_i\big)_{i=1}^{n+1}} \\
& \ge \lambda_{n+1} \bignorm[2]{\big(\hf_i\big)_{i=1}^{n+1}} = \lambda_{n+1} \norm[\cf]{f} = \lambda_{n+1} \rho.
\end{align*}
Thus, $\lambda_{n+1} \le \varepsilon/\rho$, and $n \ge \min\{ n : \lambda_{n+1} \le \varepsilon/\rho \}$.  Hence, algorithm $\hA$ defined in \eqref{optBallalg} is optimal for $\ca(\cb_{\rho})$.

We defined an algorithm $A \in \ca(\ch)$ to have \emph{essentially no worse cost} then an algorithm $A^* \in \ca(\ch^*)$ if for some number $\omega$,
\begin{gather}
\cost(A,\ch,\varepsilon) \le \cost(A^*,\ch^*,\omega\varepsilon), \qquad \forall \varepsilon > 0, \\
\cost(A,\ch,\varepsilon,\rho) \le \cost(A^*,\ch^*,\omega\varepsilon,\rho), \qquad \forall \varepsilon,\rho > 0.
\end{gather}
If the costs of two algorithms are essentially no worse than each other, then we call them essentially the same.  An algorithm whose cost is essentially no worse than the best possible algorithm, is called \emph{essentially optimal}.

@@

Our rigorous, data-driven error bound assumes the input to have steadily---but not necessarily monotonically---decaying series coefficients.  The cone of nice inputs, $\cc$, is defined in Section \ref{sec:cone}.  For such inputs, we construct an adaptive algorithm, $\tA \in \ca(\cc)$, in Section \ref{sec:adaptalgo},  where $\tA(f,\varepsilon) = A_{n^*}(f)$ for some $n^*$.  The number of series coefficients sampled, $n^*$, is adaptively determined so that $\tA(f,\varepsilon)$ satisfies the error condition in \eqref{adapErrCrit}.  The computational cost of $\tA$ is given in  Theorem \ref{thm:compcost}.  Section \ref{sec:opt} shows that our new algorithm is asymptotically optimal, i.e., ???.  Section \ref{sec:examp} provides an example of our algorithm.  Concluding remarks are give in Section \ref{sec:conc}.  We close this section with an example of the function spaces and solution operator that we consider.

\begin{exmp} Consider the case of function approximation for periodic functions defined on the [0,1], and algorithm $\hA$ defined in \eqref{optBallalg}:
\begin{align*}
f &= \sum_{k \in \integers} \hf(k) \hu_{k} (\vx) = \sum_{i=1}^n \hf_i u_i(x), 
& S(f) & = \sum_{k \in \integers} \hf(k) \hlambda_k \hv_k = \sum_{i=1}^n \hf_i \lambda_i \hv_i, \\
\hv_{k} (x) &:= \begin{cases} 1, & k = 0, \\
\displaystyle \sqrt{2} \sin(2\pi k x), & k > 0, \\
\displaystyle \sqrt{2} \cos(2\pi k x), & k < 0, \\
\end{cases} 
& v_i(x) &= \begin{cases} \hv_{-i/2}, & i \text{ even},\\
\hv_{(i-1)/2}, & i \text{ odd},
\end{cases} \\
\hlambda_{k} &:= \begin{cases} 1, & k = 0, \\
\displaystyle \frac{1}{\abs{k}^r}, & k \ne 0,
\end{cases}
& \lambda_i &= \hlambda_{\lfloor i/2\rfloor} = \frac{1}{\max(1,\lfloor i/2\rfloor)^r}, \\
\hu_{k} (x) &:= \hlambda_k\hv_k(x),
& u_i(x) &= \begin{cases} \hu_{-i/2}, & i \text{ even},\\
\hu_{(i-1)/2}, & i \text{ odd},
\end{cases}
\\
&&\hf_i& = \begin{cases} \hf(-i/2), & i \text{ even},\\
\hf((i-1)/2), & i \text{ odd},
\end{cases}
\end{align*}
\begin{align*}
\comp(\ca(\cb_{\rho}),\varepsilon) &= \cost(\hA,\cb_{\rho},\varepsilon) 
=\min\{ n : \lambda_{n+1} \le \varepsilon/\rho \} \\
& = \min \left\{ n : \frac{1}{\lfloor (n+1)/2 \rfloor^r} \le \frac{\varepsilon}{\rho} \right\} 
= 2 \left \lceil \left(\frac{\rho}{\varepsilon} \right)^{1/r} \right \rceil - 1.
% & \lambda_{n+1} = \frac{1}{\lceil (n+1)/2 \rceil^r}.
\end{align*}
Here $\cg=\cl_2[0,1]$.  The larger the non-negative parameter $r$ is, the smaller $\cb_\rho$ is, and the smaller  $\cost(\hA,\cb_{\rho},\varepsilon)$ is.  For $r = 0$, $\cost(\hA,\cb_{\rho},\varepsilon) = \infty$.
\end{exmp}

\section{Assuming a Steady Decay of the Series Coefficients of the Solution} \label{sec:cone}

@@

Using the expression for the error in \eqref{errOpt} and the definition of the cone in  \eqref{decayconedef}, one can now derive a data-driven error bound for $j \in \naturals$:
\begin{align}
\nonumber
\norm[\cg]{S(f)-A_{n_j}(f)} &= \norm[2]{\left(\lambda_{i} \hf_{i} \right)_{i = n_j+1}^\infty} = \left\{\sum_{r=1}^\infty \sum_{i=n_{j+r-1}+1}^{n_{j+r}}  \abs{\lambda_{i}\hf_{i} }^{2}  \right\}^{1/2}\\
\nonumber
&= \norm[2]{ \left(\sigma_{j+r}(f)\right)_{r=1}^{\infty}} \\
&\le \norm[2]{ \left(ab^r\sigma_{j}(f)\right)_{r=1}^{\infty}}
 = \displaystyle ab \sqrt{\frac{1}{1 - b^2}}\sigma_{j}(f)
 \label{algoineq}
\end{align}
The upper and lower error bounds depend only on the function data and the parameters defining $\cc$.  The error vanishes as $j \to \infty$ because $\sigma_j(f) \le ab^{j-1} \sigma_1(f) \to 0$ as $j \to \infty$.  Moreover, the error of $A_{n_j}(f)$ is asymptotically the no worse than $\sigma_j(f)$, whose rate of decay need not be postulated in advance. Our adaptive algorithm in the Section \ref{sec:adaptalgo} increases $j$ until the right hand side is smaller than the error tolerance.

@@

\begin{theorem}\label{thm:compcost}
The algorithm, $\tA$, defined in Algorithm \ref{algo2} lies in $\ca(\cc)$ and has computational cost $\cost(\tA,f,\varepsilon)=n_{j^*}$, where $j^*$ is defined implicitly by the inequalities 
\[
\sigma_{j^*}(f) \le \frac{\varepsilon\sqrt{1 - b^2}}{ab}  < \sigma_{j}(f), \quad 1 \le j < j^*.
\]
Moreover, $\cost(\tA,\varepsilon,\rho) \le n_{j^\dagger}$, where $j^\dagger$ satisfies the following upper bound:
\begin{equation} \label{jdagger}
j^\dagger \le \min \left \{j \in \naturals : \frac{\rho^2}{\varepsilon^2} \le \frac{(1 - b^2)}{a^2b^2} \left[ \sum_{k=1}^{j-1} \frac{b^{2(k-j)}}{a^2\lambda_{n_{k-1}+1}^2} + \frac{1}{\lambda_{n_{j-1}+1}^2}\right]   \right\}.
\end{equation}

@@

To derive an upper bound on $\cost(\tA,\varepsilon,\rho)$ we start by noting some properties of $\sigma_j(f)$ for all $f \in \cc$:
\begin{multline} \label{normsigineq}
\lambda_{n_j} \bignorm[2]{\big( \hf_i \big)_{i=n_{j-1}+1}^{n_j}} \le 
\bignorm[2]{\big( \lambda_i \hf_i \big)_{i=n_{j-1}+1}^{n_j}} = \sigma_j(f) 
\\
\le \lambda_{n_{j-1}+1} \bignorm[2]{\big( \hf_i \big)_{i=n_{j-1}+1}^{n_j}}.
\end{multline}
A rough upper bound on $\cost(\tA,\rho,\varepsilon)$ may be obtained by noting that  for any $f \in \cc$ that also lies in the ball of radius $\rho$, and for any $j < j^\dagger$, it follows by \eqref{normsigineq} and  Theorem \ref{thm:compcost} that 
\begin{equation*}
\rho \ge \norm[\cf]{f} \ge \bignorm[2]{\big( \hf_i \big)_{i=n_{j-1}+1}^{n_j}} \ge \frac{\sigma_j(f)}{\lambda_{n_{j-1}+1}} > \frac{\varepsilon\sqrt{1 - b^2}}{ab \lambda_{n_{j-1}+1}}
\end{equation*}
Thus, one upper bound on $j^\dagger$ is the smallest $j$ violating the above inequality:
\begin{equation} \label{jdaggerlast}
j^\dagger \le \min \left \{j \in \naturals :  \lambda_{n_{j-1}+1} \le \frac{\varepsilon\sqrt{1 - b^2}}{ab \rho} \right\}.
\end{equation}

>>>>

\section{Introduction}
Adaptive algorithms determine the design and sample size needed to solve problems to the desired accuracy.  A priori upper bounds on some norm of the input function are not needed, but some underlying assumptions about the input function are required for the adaptive algorithm to succeed.  Here we consider general linear problems where a finite number of series coefficients of the input function are used to obtain an approximate solution.  The proposed algorithm produces an approximation with guaranteed accuracy.  Moreover, we demonstrate that  computational cost of our algorithm is essentially no worse than that of the best possible algorithm.  The set of input functions for which our adaptive algorithm is defined is a cone.

Let $\cf$ be a separable Hilbert space of input functions with orthonormal basis $\{u_i\}_{i=1}^\infty$, let $\cg$ be a separable Hilbert space of outputs  with orthonormal basis $\{v_i\}_{i=1}^\infty$, and let their norms be defined as the $\ell^2$-norms of their series coefficients:  
\begin{subequations}\label{probDef}
\begin{gather}
f = \sum_{i=1}^\infty \hf_i u_i \in \cf, \qquad \norm[\cf]{f}=\bignorm[2]{(\hf_i\big)_{i=1}^\infty}, \\
g = \sum_{i=1}^\infty \hg_i v_i \in \cg, \qquad \norm[\cg]{g}=\bignorm[2]{(\hg_i\big)_{i=1}^\infty}.
\end{gather}
Let the bases of spaces of inputs and outputs be chosen by the singular value decomposition so that the linear solution operator, $S:\cf \to \cg$, satisfies
\begin{gather}
S(u_i) = \lambda_i v_i, \quad i \in \naturals, \qquad S(f) = \sum_{i=1}^n \lambda_i \hf_i v_i, \\
\lambda_1 \ge \lambda_2 \ge \cdots > 0, \quad \lim_{i \to \infty} \lambda_i = 0, \qquad
\norm[\cf \to \cg]{S} := \sup_{f \ne 0} \frac{\norm[\cg]{S(f)}}{\norm[\cf]{f}} = \lambda_1.
\end{gather}
\end{subequations}
This setting includes, for example, the recovery of functions, derivatives, and indefinite integrals.  We are interested in cases where the exact solution generally requires an infinite number of series coefficients, $\hf_i$, Thus, this setting is not relevant for evaluating the definite integral or any other linear functional, since in those cases $\lambda_2 = \lambda_3 = \cdots = 0$.

Let $\ch$ be any subset of $\cf$, and let $\ca(\ch)$ denote the set of deterministic algorithms that successfully approximate the solution operator $S : \ch \to \cg$ to within some error tolerance for all inputs in $\ch$:
\begin{multline} \label{adapErrCrit}
\ca(\ch) : = \left\{ \text{algorithms } A:\ch \times (0,\infty) \rightarrow \cg : 
\right . \\ \left .
\bignorm[\cg]{S(f) - A(f,\varepsilon)} \le \varepsilon \ \forall f \in \ch, \ \varepsilon > 0 
\right \}.
\end{multline}
Algorithms in $\ca(\ch)$ are allowed to sample adaptively any bounded, linear functionals
of the input function.  They must require only a finite number of linear functionals for each input function and positive tolerance.  The definition of $\ch$ can be used to construct algorithms in $\ca(\ch)$, but no other a priori knowledge about the input functions is available.  Following \cite{KunEtal19a} we call a problem \emph{solvable} for inputs $\ch$ if $\ca(\ch)$ is non-empty.

Our problem is not solvable for $\cf$, as can be demonstrated by contradiction. For any potential algorithm, we show that there exists some $f \in \cf$, that looks like $0$ to the algorithm, but for which $S(f)$ is far from $S(0) = 0$.   Choose any $A \in \ca(\cf)$ and $\varepsilon > 0$, and let $L_1, \ldots, L_n$ be the linear functionals are used to compute $A(0,\varepsilon)$. Since the output space, $\cg$, is infinite dimensional and $n$ is finite, there exists some nonzero $f \in \cf$ satisfying that $L_1(f) = \cdots = L_n(f) = 0$ with non-zero $S(f)$. This means that $A(cf,\varepsilon) = A(0,\varepsilon)$ for any real $c$.  Since $A$ successfully produces an approximation with error no greater than $\varepsilon$ for both $0$ and $cf$,
\begin{align*}
    \varepsilon &\ge \frac 12 \left[ \norm[\cg]{S(0) - A(0,\varepsilon)} + \norm[\cg]{S(cf) - A(cf,\varepsilon)}\right] \\
    & =  \frac 12 \left[ \norm[\cg]{0 - A(0,\varepsilon)} + \norm[\cg]{S(cf) - A(0,\varepsilon)}\right] \\
    & \ge  \frac {\abs{c}\norm[\cg]{S(f)}}2  \qquad \text{by the triangle inequality}.
\end{align*}
Since $S(f) \ne 0$, it is impossible for this inequality to hold for all real $c$.  The presumed $A$ does not exist, $\ca(\cf)$ is empty, and our problem is not solvable for $\cf$. However, it is solvable for well-chosen subsets of $\cf$, as will be shown in the sections below.

The computational cost of an algorithm $A \in \ca(\ch)$ for $f \in \ch$ and error tolerance $\varepsilon$ is denoted $\cost(A,f,\varepsilon)$, and is defined as the number of linear functional values required to produce $A(f,\varepsilon)$.  By overloading the notation, we define the cost of algorithms as 
\begin{equation*}
\cost(A,\ch,\varepsilon) : = \sup \{\cost(A,f,\varepsilon) : f \in \ch \} \qquad \forall \varepsilon > 0.
\end{equation*}
For unbounded sets, $\ch$, this cost may be infinite.  Therefore, it is also helpful to define the cost of algorithms for input functions in $\ch \cap \cb_{\rho}$, where $\cb_{\rho} : = \{ f \in \cf : \norm[\cf]{f} \le \rho \}$ is the ball of radius $\rho$:
\begin{equation*}
\cost(A,\ch,\varepsilon,\rho) : = \sup \{\cost(A,f,\varepsilon) : f \in \ch \cap \cb_{\rho} \} \qquad \forall \rho > 0, \ \varepsilon > 0.
\end{equation*}
Finally, we define the complexity of the problem as computational cost of the best algorithm:
\begin{gather*}
\comp(\ca(\ch),\varepsilon) := \min_{A \in \ca(\ch)} \cost(A, \ch, \varepsilon), \\
\comp(\ca(\ch),\varepsilon,\rho) := \min_{A \in \ca(\ch)} \cost(A, \ch, \varepsilon, \rho).
\end{gather*}
Note that $\comp(\ca(\ch),\varepsilon,\rho) \ge \comp(\ca(\ch \cap \cb_{\rho}),\varepsilon)$.  In the former case, the algorithm is unaware  that the input function has norm no greater than $\rho$.

@@

To prove this, let $A^*$ be an arbitrary algorithm in $\ca(\cb_{\rho})$, and let $L_1, \ldots, L_N$ be the linear functionals chosen when evaluating this algorithm for the zero function with tolerance $\varepsilon$.  Thus, $A^*(0,\varepsilon)$ is some function of $(L_1(0) , \ldots, L_N(0)) = (0, \ldots, 0)$.  Let $f$ be a linear combination of $u_1, \ldots, u_{n+1}$ with norm $\rho$ satisfying  $L_1(f) = \cdots = L_N(f) = 0$, then $A^*(\pm f) = A^*(0)$, and
\begin{align*}
\varepsilon & \ge \max_{\pm} \norm[\cg]{S(\pm f) - A^*(\pm f)} =  \max_{\pm} \norm[\cg]{\pm S( f) - A^*(0)} \\
& \ge \frac 12 \left [ \norm[\cg]{S( f) - A^*(0)} + \norm[\cg]{- S( f) - A^*(0)}\right] \\
& \ge \norm[\cg]{S(f)} 
= \bignorm[2]{\big(\lambda_i\hf_i\big)_{i=1}^{N+1}} \\
& \ge \lambda_{N+1} \bignorm[2]{\big(\hf_i\big)_{i=1}^{N+1}} = \lambda_{N+1} \norm[\cf]{f} = \lambda_{N+1} \rho.
\end{align*}
Thus, $\lambda_{N+1} \le \varepsilon/\rho$, and $N \ge \min\{ n : \lambda_{n+1} \le \varepsilon/\rho \} = \cost(\tA,\cb_\rho,\varepsilon)$.  Hence, algorithm $\hA$ defined in \eqref{optBallalg} is optimal for $\ca(\cb_{\rho})$.

Our goal is not only to construct algorithms in $\ca(\ch)$ for some $\ch$, but also to determine whether the computational cost of our algorithms are reasonable.  We define an algorithm $A \in \ca(\ch)$ to have \emph{essentially no worse cost} then an algorithm $A^* \in \ca(\ch^*)$ if for some number $\omega$,
\begin{gather}
\cost(A,\ch,\varepsilon) \le \cost(A^*,\ch^*,\omega\varepsilon), \qquad \forall \varepsilon > 0, \\
\cost(A,\ch,\varepsilon,\rho) \le \cost(A^*,\ch^*,\omega\varepsilon,\rho), \qquad \forall \varepsilon,\rho > 0.
\end{gather}
If the costs of two algorithms are essentially no worse than each other, then we call them essentially the same.  An algorithm whose cost is essentially no worse than the best possible algorithm, is called \emph{essentially optimal}.

@@

Our rigorous, data-driven error bound assumes the input to have steadily---but not necessarily monotonically---decaying series coefficients.  The cone of nice inputs, $\cc$, is defined in Section \ref{sec:cone}.  For such inputs, we construct an adaptive algorithm, $\tA \in \ca(\cc)$, in Section \ref{sec:adaptalgo},  where $\tA(f,\varepsilon) = A_{n^*}(f)$ for some $n^*$.  The number of series coefficients sampled, $n^*$, is adaptively determined so that $\tA(f,\varepsilon)$ satisfies the error condition in \eqref{adapErrCrit}.  The computational cost of $\tA$ is given in  Theorem \ref{thm:compcost}.  Section \ref{sec:opt} shows that our new algorithm is essentially optimal (see Theorem \ref{thm:CostNoWorse}).  Section \ref{sec:examp} provides an example of our algorithm.  Concluding remarks are give in Section \ref{sec:conc}.  We close this section with an example of the function spaces and solution operator that we consider.

\begin{exmp} Consider the case of function approximation for periodic functions defined on the [0,1], and algorithm $\hA$ defined in \eqref{optBallalg}:
\begin{align*}
f &= \sum_{k \in \integers} \hf(k) \hu_{k}  = \sum_{i=1}^n \hf_i u_i, 
& S(f) & = \sum_{k \in \integers} \hf(k) \hlambda_k \hv_k = \sum_{i=1}^n \hf_i \lambda_i \hv_i, \\
\hv_{k} (x) &:= \begin{cases} 1, & k = 0, \\
\displaystyle \sqrt{2} \sin(2\pi k x), & k > 0, \\
\displaystyle \sqrt{2} \cos(2\pi k x), & k < 0, \\
\end{cases} 
& v_i(x) &= \begin{cases} \hv_{-i/2}, & i \text{ even},\\
\hv_{(i-1)/2}, & i \text{ odd},
\end{cases} \\
\hlambda_{k} &:= \begin{cases} 1, & k = 0, \\
\displaystyle \frac{1}{\abs{k}^r}, & k \ne 0,
\end{cases}
& \lambda_i &= \hlambda_{\lfloor i/2\rfloor} = \frac{1}{\max(1,\lfloor i/2\rfloor)^r}, \\
\hu_{k} &:=  \hlambda_k \hv_k,
& u_i & \lambda_i v_i = \begin{cases} \hu_{-i/2}, & i \text{ even},\\
\hu_{(i-1)/2}, & i \text{ odd},
\end{cases}
\\
&&\hf_i& = \begin{cases} \hf(-i/2), & i \text{ even},\\
\hf((i-1)/2), & i \text{ odd},
\end{cases}
\end{align*}
\begin{align*}
\comp(\ca(\cb_{\rho}),\varepsilon) &= \cost(\hA,\cb_{\rho},\varepsilon) 
=\min\{ n : \lambda_{n+1} \le \varepsilon/\rho \} \\
& = \min \left\{ n : \frac{1}{\lfloor (n+1)/2 \rfloor^r} \le \frac{\varepsilon}{\rho} \right\} 
= 2 \left \lceil \left(\frac{\rho}{\varepsilon} \right)^{1/r} \right \rceil - 1.
% & \lambda_{n+1} = \frac{1}{\lceil (n+1)/2 \rceil^r}.
\end{align*}
Here $\cg=\cl_2[0,1]$.  The larger the non-negative parameter $r$ is, the faster the $\lambda_i$ tend to 0 as $ i \to 0$, the smaller $\cb_\rho$ is, and the smaller  $\cost(\hA,\cb_{\rho},\varepsilon)$ is.  For $r = 0$, $\cost(\hA,\cb_{\rho},\varepsilon) = \infty$.
\end{exmp}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Assuming a Steady Decay of the Series Coefficients of the Solution} \label{sec:cone}

@@

From the expression for the error in \eqref{errOpt} and the definition of the cone in  \eqref{decayconedef}, one can now derive a data-driven error bound for $j \in \naturals$:
\begin{align}
\nonumber
\norm[\cg]{S(f)-A_{n_j}(f)} &= \norm[2]{\left(\lambda_{i} \hf_{i} \right)_{i = n_j+1}^\infty} = \left\{\sum_{r=1}^\infty \sum_{i=n_{j+r-1}+1}^{n_{j+r}}  \abs{\lambda_{i}\hf_{i} }^{2}  \right\}^{1/2}\\
\nonumber
&= \norm[2]{ \bigl(\sigma_{j+r}(f)\bigr)_{r=1}^{\infty}} \\
&\le \norm[2]{ \bigl(ab^r\sigma_{j}(f)\bigr)_{r=1}^{\infty}}
 = \displaystyle ab \sqrt{\frac{1}{1 - b^2}}\sigma_{j}(f)
 \label{algoineq}
\end{align}
The upper and lower error bounds depend only on the function data and the parameters defining $\cc$.  The error vanishes as $j \to \infty$ because $\sigma_j(f) \le ab^{j-1} \sigma_1(f) \to 0$ as $j \to \infty$.  Moreover, the error of $A_{n_j}(f)$ is asymptotically the no worse than $\sigma_j(f)$, whose rate of decay need not be postulated in advance. Our adaptive algorithm in the Section \ref{sec:adaptalgo} increases $j$ until the right hand side is smaller than the error tolerance.

@@

\begin{theorem}\label{thm:compcost}
The algorithm, $\tA$, defined in Algorithm \ref{algo2} lies in $\ca(\cc)$ and has computational cost $\cost(\tA,f,\varepsilon)=n_{j^*}$, where $j^*$ is defined implicitly by the inequalities 
\begin{equation} \label{eq:OurAlgjstar}
j^* = \min\left \{ j \in \naturals : \sigma_{j}(f) \le \frac{\varepsilon\sqrt{1 - b^2}}{ab}  \right\}.
\end{equation}
Moreover, $\cost(\tA,\varepsilon,\rho) \le n_{j^\dagger}$, where $j^\dagger$ satisfies the following upper bound:
\begin{equation} \label{jdagger}
j^\dagger \le \min \left \{j \in \naturals : \frac{\rho^2}{\varepsilon^2} \le \frac{(1 - b^2)}{a^2b^2} \left[ \sum_{k=1}^{j-1} \frac{b^{2(k-j)}}{a^2\lambda_{n_{k-1}+1}^2} + \frac{1}{\lambda_{n_{j-1}+1}^2}\right]   \right\}.
\end{equation}

@@

For the remainder of the proof consider $\rho$ and $\varepsilon$ fixed.  To derive an upper bound on $n_{j^\dagger} = \cost(\tA,\varepsilon,\rho)$ we first note some properties of $\sigma_j(f)$ for all $f \in \cc$:
\begin{multline} \label{normsigineq}
\lambda_{n_j} \bignorm[2]{\big( \hf_i \big)_{i=n_{j-1}+1}^{n_j}} \le 
\bignorm[2]{\big( \lambda_i \hf_i \big)_{i=n_{j-1}+1}^{n_j}} = \sigma_j(f) 
\\
\le \lambda_{n_{j-1}+1} \bignorm[2]{\big( \hf_i \big)_{i=n_{j-1}+1}^{n_j}}.
\end{multline}

A rough upper bound on $j^\dagger$ may be obtained by noting that for any $f \in \cc \cap \cb_\rho$ and for any $j < j^* \le j^\dagger$, it follows from \eqref{eq:OurAlgjstar} and \eqref{normsigineq} that 
\begin{equation*}
\rho \ge \norm[\cf]{f} \ge \bignorm[2]{\big( \hf_i \big)_{i=n_{j-1}+1}^{n_j}} \ge \frac{\sigma_j(f)}{\lambda_{n_{j-1}+1}} > \frac{\varepsilon\sqrt{1 - b^2}}{ab \lambda_{n_{j-1}+1}}
\end{equation*}
Thus, one upper bound on $j^\dagger$ is the smallest $j$ violating the above inequality:
\begin{equation} \label{jdaggerlast}
j^\dagger \le \min \left \{j \in \naturals :  \lambda_{n_{j-1}+1} \le \frac{\varepsilon\sqrt{1 - b^2}}{ab \rho} \right\}.
\end{equation}

The tighter upper bound in Theorem \ref{thm:compcost} may be obtained by a more careful argument in a similar vein.  
For any $f \in  \cc \cap \cb_\rho$ and for any $j < j^* \le j^\dagger$,
\begin{align*}
\rho^2 &\ge \norm[\cf]{f}^2 = \bignorm[2]{\big( \hf_i \big)_{i=1}^{\infty}}^2 \\
& \ge \sum_{k=1}^j \bignorm[2]{\big( \hf_i \big)_{i=n_{k-1}+1}^{n_k}}^2 \qquad \forall j \ge 1\\
& \ge \sum_{k=1}^j \frac{\sigma_k^2(f)}{\lambda_{n_{k-1}+1}^2} \qquad \text{by \eqref{normsigineq}}\\
& \ge \sum_{k=1}^{j-1} \frac{ b^{2(k-j)}\sigma_j^2(f)}{a^{2}\lambda_{n_{k-1}+1}^2} + \frac{\sigma_j^2(f)}{\lambda_{n_{j-1}+1}^2} \qquad \text{by \eqref{decayconedef}} \\
& = \sigma_j^2(f) \left[ \sum_{k=1}^{j-1} \frac{b^{2(k-j)}}{a^{2} \lambda_{n_{k-1}+1}^2} + \frac{1}{\lambda_{n_{j-1}+1}^2}\right].
\end{align*}
Note that the quantity in the square brackets is an increasing function of $j$ because as $j$ increases, the sum becomes includes more terms and $b^{2(k-j)}$ also increases.

For all $j < j^* \le j^\dagger$ it follows from \eqref{eq:OurAlgjstar} that
\begin{equation*}
\rho^2 > \frac{\varepsilon^2(1 - b^2)}{a^2b^2} \left[ \sum_{k=1}^{j-1} \frac{ b^{2(k-j)}}{a^{2}\lambda_{n_{k-1}+1}^2} + \frac{1}{\lambda_{n_{j-1}+1}^2}\right].
\end{equation*}
Thus, any $j$ that violates the above inequality, must satisfy $j \ge j^\dagger$, establishing \eqref{jdagger}.
\end{proof}
